---
title: "Tidying Data Assignment"
author: ''
date: ''
---
# Please complete all the tasks listed below. After completing the tasks download the .Rmd file and upload in the peer review item for grading.

# Additionally please write text between the code chunks explaining what each code chunk is about.

# Refer the linked online textbook in case of any issues.

Load the tidyverse library
```{r}
library(tidyverse)
```

##### Question 1. 

The built in billboard dataset is not tidy. Describe why it is not tidy and then tidy the dataset.

The billboard dataset is not tidy because:

- Each song is represented in multiple rows with weekly rankings as separate columns.
- Multiple observations are stored in the same table, making it difficult to manipulate and analyze.

```{r question-1-response}

# First gather up all the week entries into a row for each week for each song (where there is an entry)
# Load billboard dataset
data("billboard")

# Then, convert the week variable to a number and figure out the date corresponding to each week on the chart
# Tidy the dataset
billboard_tidy <- billboard %>%
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    names_prefix = "wk",
    values_to = "rank",
    values_drop_na = TRUE
  ) %>%
  mutate(week = as.integer(week))

# Sort the data by artist, track and week. Here are what your first entries should be (formatting can be different):
#>  A tibble: 5,307 x 5
#   artist  track                   date.entered  week  rank   date
 #    <chr>   <chr>                   <date>       <int> <dbl>   <date>
 #  1 2 Pac   Baby Don't Cry (Keep... 2000-02-26       1    87 2000-02-26
 #  2 2 Pac   Baby Don't Cry (Keep... 2000-02-26       2    82 2000-03-04
 #  3 2 Pac   Baby Don't Cry (Keep... 2000-02-26       3    72 2000-03-11
 #  4 2 Pac   Baby Don't Cry (Keep... 2000-02-26       4    77 2000-03-18
 #  5 2 Pac   Baby Don't Cry (Keep... 2000-02-26       5    87 2000-03-25
 #  6 2 Pac   Baby Don't Cry (Keep... 2000-02-26       6    94 2000-04-01
 #  7 2 Pac   Baby Don't Cry (Keep... 2000-02-26       7    99 2000-04-08
 #  8 2Ge+her The Hardest Part Of ... 2000-09-02       1    91 2000-09-02
 #  9 2Ge+her The Hardest Part Of ... 2000-09-02       2    87 2000-09-09
 # 10 2Ge+her The Hardest Part Of ... 2000-09-02       3    92 2000-09-16
 # … with 5,297 more rows

# Display the tidy dataset
head(billboard_tidy)

```

##### Question 2. 

Tidy the "fish_encounters" dataset of fish spotting by monitoring stations. Make the NA into 0 using the option "values_fill = list(seen = 0)"


```{r question-2-response}

# Load fish_encounters dataset
data("fish_encounters")

# str(fish_encounters)

# Tidy the dataset with NA filled with 0
fish_encounters_tidy <- fish_encounters %>%
  complete(fish, station) %>%  # Ensure all combinations of fish and station are present
  replace_na(list(seen = 0))  # Replace NA values with 0


# Display the tidy dataset
head(fish_encounters_tidy)

```

##### Question 3. 

Import the flowers1 dataset. Tidy and pivot the data. Hint: use "read_csv2()" to read in the dataset 

```{r question-3-response}

# Import flowers1 dataset
flowers1 <- read_csv2("flowers1.csv")

str(flowers1)

spec(flowers1)

# Pivot longer
flowers1_long <- flowers1_tidy %>%
  pivot_longer(
    cols = -c(Time, replication),
    names_to = "Variable",
    values_to = "Value"
  )

# Display the tidy dataset
head(flowers1_tidy)

```

##### Question 4.

Import the flowers2 dataset. Tidy the dataset by turning the one column into 3 separate columns

```{r question-4-response}

# Import flowers2 dataset
flowers2 <- read_csv2("flowers2.csv")

# str(flowers2)

# Separate the Flowers/Intensity column into two columns: Flowers and Intensity
flowers2_tidy <- flowers2 %>%
  separate(
    col = `Flowers/Intensity`,
    into = c("Flowers", "Intensity"),
    sep = "/"
  ) %>%
  mutate(
    Flowers = as.numeric(Flowers),
    Intensity = as.numeric(Intensity)
  )

# Display the tidy dataset
head(flowers2_tidy)


```

##### Question 5.

In the following dataset, turn the implicit missing values to explicit 

```{r question-5-response}
output <- tibble(
      treatment   = c("a", "b", "a", "c", "b"),
      gender   = factor(c("M", "F", "F", "M", "M"), levels = c("M", "F", "O")),
      return = c(1.5, 0.75,  0.5,  1.8,  NA)
    )
output

# Make implicit missing values explicit
output_explicit <- output %>%
  complete(treatment, gender)

# Display the dataset with explicit missing values
output_explicit


```


#####Question 6.

Import the weather dataset as weather. Use "pivot_longer()" to put the days all in one column, then use "pivot_wider" to separate tmax and tmin into separate columns. Print the summary of the final resulting dataset 

```{r question-6-response}
# Import weather dataset
weather <- read_csv("weather.csv")

# str(weather)

# Tidy the dataset
weather_tidy <- weather %>%
  pivot_longer(
    cols = starts_with("d"),
    names_to = "day",
    names_prefix = "d",
    values_to = "value"
  ) %>%
  pivot_wider(
    names_from = element,
    values_from = value
  )

# Print summary of the final dataset
summary(weather_tidy)

```

###### Question 7. 

Load the built in "anscombe" data frame and use "pivot_longer()" to separate all the x and y columns and categorize them into 4 sets

```{r question-7-response}
# Load anscombe dataset
data("anscombe")

# Tidy the dataset
anscombe_tidy <- anscombe %>%
  pivot_longer(
    cols = everything(),
    names_to = c(".value", "set"),
    names_pattern = "(.)(.)"
  ) %>%
  mutate(set = as.integer(set))

# Display the tidy dataset
head(anscombe_tidy)
```
